---
title: "Crowding in Grampian"
output:
  html_document: default
---


```{r, include = F}
library(here)
library(tidyverse)
library(janitor)
library(sf)
library(mapview)
library(gt)
library(cowplot)
library(plotly)
library(ggrepel)

theme_set(theme_cowplot())
```


```{r, include = F}
#load SIMD data files

simd_zones <- 
  read_csv(here("data", "SIMD+2020v2+-+datazone+lookup.csv")) 

simd_indicators <- 
  read_csv(here("data", "SIMD2020v2_indicators.csv")) %>%
   mutate(across(everything(), ~ str_remove_all(., "%")),
          across(everything(), ~ str_remove_all(., "\\*")),
          across(Total_population:nocentralheat_rate, ~ as.numeric(.)))

postcodes <- read_csv(here("data", "SIMD+2020v2+-+postcode+lookup.csv")) 

#load shapefiles for maps

datazone_sf <- st_read(here("data", "sc_dz_11.shp"))

#load data from 2011 census bulk files
#clean variable names to remove caps and spaces

ethnicity <- read_csv(here("data", "SNS Data Zone 2011 blk", "KS201SC.csv")) %>% 
  clean_names()

english_skills <- read_csv(here("data", "SNS Data Zone 2011 blk", "QS213SC.csv")) %>%
  clean_names()

country_birth <- read_csv(here("data", "SNS Data Zone 2011 blk", "KS204SC.csv")) %>%
  clean_names() 

```

```{r, include = F}

#join deprivation to census files

simd <- left_join(simd_indicators, simd_zones, by = c("Data_Zone" = "DZ")) %>%
  left_join(., english_skills, by = c("Data_Zone" = "x1")) %>%
  left_join(., ethnicity, by = c("Data_Zone" = "x1")) %>%
  left_join(., country_birth, by = c("Data_Zone" = "x1"))

#calculate percentages from census data

simd <-
  mutate(simd, 
         perc_not_born_uk_roi = 100- ((england + scotland + wales + northern_ireland + republic_of_ireland)/all_people.y * 100),
         perc_not_working_age = 100- (Working_age_population/Total_population *100),
         perc_polish = white_polish/all_people.x * 100,
         perc_non_white = 100 - (white/all_people.x * 100)
         )

#combine all postcodes in each datazone
postcodes_in_datazone <-
postcodes %>%
  group_by(DZ) %>%
  summarise(postcodes = toString(Postcode)) 

```


```{r, include = F}
#select local authorities' data 
#calculate population density
#create percentile variable and flag the 95th+
#join to shapefile
#(this is repeated 3x and should be a function)

aberdeen <-
  simd %>%
  filter(Council_area == "Aberdeen City") 

aberdeen_data_zones <- pull(aberdeen, Data_Zone)

aberdeen_sf <- filter(datazone_sf, DataZone %in% aberdeen_data_zones)

aberdeen_sf <-
  merge(aberdeen_sf, aberdeen, by.x = "DataZone", by.y = "Data_Zone")

aberdeen_sf <-
  aberdeen_sf %>% 
  mutate(pop_per_km2 = Total_population/StdAreaKm2,
         pop_density_percentile = ntile(pop_per_km2, 100),
         overcrowded_rate_percentile = ntile(overcrowded_rate, 100),
         very_crowded = 
           ifelse(
             pop_density_percentile >= 95 | overcrowded_rate_percentile >= 95, 1, 0))
   
aberdeen <-  
  select(aberdeen_sf, 
         DataZone, pop_per_km2, pop_density_percentile, overcrowded_rate_percentile) %>%
  left_join(aberdeen, ., by = c("Data_Zone" = "DataZone"))

shire <-
  simd %>%
  filter(Council_area == "Aberdeenshire") 

shire_data_zones <- pull(shire, Data_Zone)

shire_sf <- filter(datazone_sf, DataZone %in% shire_data_zones)

shire_sf <-
  merge(shire_sf, shire, by.x = "DataZone", by.y = "Data_Zone")

shire_sf <-
  shire_sf %>% 
  mutate(pop_per_km2 = Total_population/StdAreaKm2,
         pop_density_percentile = ntile(pop_per_km2, 100),
         overcrowded_rate_percentile = ntile(overcrowded_rate, 100),
         very_crowded = 
           ifelse(
             pop_density_percentile >= 95 | overcrowded_rate_percentile >= 95, 1, 0)) 

shire <-  
  select(shire_sf, 
         DataZone, pop_per_km2, pop_density_percentile, overcrowded_rate_percentile) %>%
  left_join(shire, ., by = c("Data_Zone" = "DataZone"))

moray <-
  simd %>%
  filter(Council_area == "Moray") 

moray_data_zones <- pull(moray, Data_Zone)

moray_sf <- filter(datazone_sf, DataZone %in% moray_data_zones)

moray_sf <-
  merge(moray_sf, moray, by.x = "DataZone", by.y = "Data_Zone")

moray_sf <-
  moray_sf %>% 
  mutate(pop_per_km2 = Total_population/StdAreaKm2,
         pop_density_percentile = ntile(pop_per_km2, 100),
         overcrowded_rate_percentile = ntile(overcrowded_rate, 100),
         very_crowded = 
           ifelse(
             pop_density_percentile >= 95 | overcrowded_rate_percentile >= 95, 1, 0)) 

moray <-  
  select(moray_sf, 
         DataZone, pop_per_km2, pop_density_percentile, overcrowded_rate_percentile) %>%
  left_join(moray, ., by = c("Data_Zone" = "DataZone"))
```

<br/><br/>

### Background  

Scotland is divided into 7000 Data Zones of about 500-1000 people each.  
This is a short report about which zones in Grampian have:  
1) many people living in crowded homes, or  
2) many people living in a crowded neighbourhood,  
or both.

<br/><br/>

### Finding neighbourhoods with dense populations and crowded homes  

The two figures below show all the Data Zones for Aberdeen City mapped for their percent of people living in overcrowded homes versus the datazone’s total density (population/km2).  
  
Datazones are red if they are in the 95th percentile or above on either measure. The second figure is the same as the first but with the red points labeled.  

```{r, echo = F, fig.width=8, fig.height=6}

#overcrowding rate vs population density graph

aberdeen_plot <- 
  ggplot(data = aberdeen_sf, 
         aes(x = pop_per_km2, y = overcrowded_rate, label = DZname)) +
  geom_point(aes(color = factor(very_crowded)), size = 3) +
  background_grid() +
  theme(legend.position = "none") +
  scale_color_manual(values=c("grey40", "red2")) +
  ylim(0, 45) +
  labs(
    x = "\nPeople per km2", 
    y = "People living in overcrowded homes (%)\n",
    subtitle = "Population density measures for Aberdeen City neighbourhoods (zones of ca. 750 people each)\n", 
    caption= "red = 95th+ percentile")

aberdeen_plot
```

  
```{r, echo = FALSE, error=F, fig.width=8, fig.height=6}

#same plot, with labels
aberdeen_plot + 
  geom_label_repel(aes(label = ifelse(very_crowded == 1, as.character(DZname), "")),
                  box.padding   = 0.35,
                  point.padding = 0.5,
                  segment.color = 'grey50',
                  force = 20,
                  max.iter = 20000) # this is a bit silly setting
```

<br/><br/>

### Mapping the most crowded  
This map shows data zones where either the overcrowding rate, the population per km2, or both is 95th percentile or above.  
These areas are the same as those in red the the figures above.  
Zoom/click for more information on each zone.  

```{r, error = F, echo = F}
#interactive map of the datazones flagged above as most crowded
aberdeen_sf %>%
  select(DataZone, Name, Total_population, SIMD2020v2_Decile, very_crowded) %>%
  mapview(
    map.types = "OpenStreetMap",
    zcol = "very_crowded",
    label = aberdeen_sf$Name,
    legend = F)
```

<br/><br/>

### Ranking the most crowded  
Data Zones with most people living in overcrowded homes or most people per km2 (95th+ percentile)  

```{r, echo = F}
#html table that fails to knit to Word doc
#same metrics as plots above
#need fixed so export to RTF in the meantime

aberdeen %>%
  select(DZname, Total_population, overcrowded_rate, pop_per_km2, overcrowded_rate_percentile, pop_density_percentile) %>%
   filter(overcrowded_rate_percentile >= 95 | pop_density_percentile >= 95) %>%
   arrange(desc(overcrowded_rate_percentile + pop_density_percentile)) %>%
  gt() %>%
 fmt_number(columns = vars(pop_per_km2), decimals = 0) %>%
  tab_header(
    title = "", 
    subtitle = "Top 5% of Data Zones with people living in overcrowded homes or most people per km2") %>%
  cols_label(
    DZname = "Data Zone",
    Total_population = "Population", 
    overcrowded_rate = "People living overcrowded (%)",
    pop_per_km2 = "Population/km2",
    overcrowded_rate_percentile = "Overcrowded percentile",
    pop_density_percentile = "Pop/km2 percentile") %>%
  tab_options(data_row.padding = px(1))
  #gtsave(filename = "table.rtf")
  #tab_options(container.height = 400)
```

<br/><br/>

### Characteristics of the most crowded  
Data Zones with most people living in overcrowded homes or most people per km2 (95th+ percentile)  

```{r, echo = F}
#html table that fails to knit to Word doc
#showing demographic characteristics
#need fixed so export to RTF in the meantime

aberdeen %>%
  filter(overcrowded_rate_percentile >= 95 | pop_density_percentile >= 95) %>%
  arrange(desc(overcrowded_rate_percentile + pop_density_percentile)) %>%
  select(DZname, Total_population, perc_not_born_uk_roi, perc_non_white, perc_polish, perc_not_working_age) %>%
  gt() %>%
  fmt_number(columns = vars(perc_not_working_age, perc_not_born_uk_roi, perc_non_white, perc_polish), decimals = 0) %>%
  tab_header(
    title = "", 
    subtitle = "Top 5% of Data Zones with people living in overcrowded homes or most people per km2") %>%
  cols_label(
    DZname = "Data Zone",
    Total_population = "Population",
    perc_not_born_uk_roi = "% Not born UK/ROI",
    perc_non_white = "% BAME",
    perc_polish = "% Polish",
    perc_not_working_age = "% Not working age",) %>%
  tab_options(data_row.padding = px(1))
  #gtsave(filename = "table2.rtf")
  #tab_options(container.height = 400)
```


```{r, echo = F}
#write most crowded areas' postcodes to file
write_csv(
filter(aberdeen, overcrowded_rate_percentile >= 95 | pop_density_percentile >= 95) %>%
  select(Data_Zone, DZname) %>%
  left_join(., postcodes_in_datazone, by = c("Data_Zone" = "DZ")),
"postcode table.csv")
```

<br/><br/>

### Sources  

Data Zone area (km2) and location and were taken from here: 
https://data.gov.uk/dataset/ab9f1f20-3b7f-4efa-9bd2-239acf63b540/data-zone-boundaries-2011

Total population, working age population, overcrowded rate and postcode mapping were taken from the Scottish Index of Multiple Deprivation 2020 version 2 
https://www.gov.scot/collections/scottish-index-of-multiple-deprivation-2020/

Country of birth and ethnicity were taken from the 2011 Census 
https://www.scotlandscensus.gov.uk/ods-web/data-warehouse.html


Code and data at https://github.com/JessButler/crowding

Contact jessicabutler@abdn.ac.uk


